Reviving stochasticity: SMBH-binary eccentricity is inherently random

### 2023-05-22 papers/paper-eccentricity/sections/discuss.tex

First, the numerical simulations performed with \ketju{} are deterministic for a given set of initial conditions, given the same computing architecture and numerical parameters (e.g. the number of computing cores used).
The scatter we observe in $\theta_\mathrm{defl}$ is thus a manifestation of Brownian motion induced by the phase space discretisation and other numerical errors, such as floating point round-off errors, denoted $\sigma_\mathrm{sim}$. 
Thus, the numerical simulations present an instance where there is zero uncertainty in the initial conditions of the system, however there is error in the numerical evolution of the system.

Second, in the real Universe there is no numerical error affecting the system evolution.
There is however uncertainty in the initial conditions of the system, whereby the chance of two galaxy merger events having the exact same initial configuration is practically zero.
We denote this uncertainty as $\sigma_\mathrm{true}$.
Intuitively, we expect that $\sigma_\mathrm{sim} > \sigma_\mathrm{real}$: in other words, $\sigma_\mathrm{sim}$ provides an upper estimate on the uncertainty we can expect from galaxy mergers in the real Universe. 
In particular, processes that perturb the impact parameter associated with the hard scattering of a SMBH pair will manifest as an effect in the resulting binary eccentricity.

>>>>

First, the numerical simulations performed with \ketju{} are deterministic for a given set of initial conditions, given the same computing architecture and numerical parameters (e.g. the number of computing cores used).
The scatter we observe in $\theta_\mathrm{defl}$ is thus a manifestation of Brownian motion induced by the phase space discretisation and other numerical errors, such as floating point round-off errors, denoted $\sigma_\mathrm{sim}$. 
Thus, the numerical simulations present an instance where there is zero uncertainty in the initial conditions of the system, however there is error in the numerical evolution of the system.
We determine the positional displacement between corresponding SMBHs across the simulations at a given mass resolution for the $e_0=0.90$ and $e_0=0.99$ orbits by moving the SMBH binary system to the centre of mass frame, and setting the SMBH positions to coincide with the origin at the beginning of the simulation. 
Computing the standard deviation in the positional displacements for corresponding SMBHs $\sigma_\mathrm{displ}$ between simulation $i$ and simulation $j\neq i$ for all times $t<t_\mathrm{bound}$, we find the SMBH positions vary of the order $\sim10\,\mathrm{pc}$ at the time of the hard scattering, which translates to an uncertainty in the deflection angle.

Second, in the real Universe there is no numerical error affecting the system evolution.
There is however uncertainty in the initial conditions of the system, whereby the chance of two galaxy merger events having the exact same initial configuration is practically zero.
We denote this uncertainty as $\sigma_\mathrm{true}$.
For the mergers of massive galaxies of the types modelled in this work, an uncertainty in the initial state of $\sigma_\mathrm{true} \sim \sigma_\mathrm{sim}$ is not unexpected, given the influence radius of a $10^8\,\Msun$ SMBH in a stellar background with velocity dispersion of the order $200\,\mathrm{km}\,\mathrm{s}^{-1}$ is $\sim 10\,\mathrm{pc}$ \drafting{From MBW, find better reference}.

Thus, any process that perturbs the impact parameter associated with the hard scattering angle will result in an eccentricity sampled from the $\theta_\mathrm{defl}$--$e_\mathrm{h}$ curve, subject to some small scatter in $e$ resulting from the hierarchical interactions of distant perturbers, or by exchanges in angular momentum with the stellar cluster immediately surrounding the SMBH. 
Critically however, the expected value of eccentricity is a direct consequence of the hard scattering angle, and is not an independent random process, and can be reliably predicted.
We caution that there is not necessarily a one-to-one mapping of the initial merger eccentricity $e_0$ and the hard binary eccentricity $e_\mathrm{h}$, unless the $\theta_\mathrm{defl}$--$e_\mathrm{h}$ curve is uniform.
As such a uniform function is not associated with the radial ($e_0>0.99$) mergers expected to occur in reality, we expect that for a given initial merger orbit $e_0$ a range of possible $e_\mathrm{h}$ would be observed.
The converse also holds: inference about merging SMBH binary eccentricity does not necessarily provide a mapping back to the initial merger eccentricity unless the history of the system is known, which is unable to be achieved in a practical sense in reality.

### 2023-05-22 papers/paper-eccentricity/sections/results.tex

After specifying the initial position and velocity, as well as the other parameters of the model,
the equation of motion can be solved using standard numerical integration techniques.
We use the error controlled 8th order Runge--Kutta method DOP853 included in the SciPy library \citep{virtanen2020}
for this purpose.
To mimic the nearly linearly plunging orbit in the high-eccentricity merger orbit simulations,
we specify the initial conditions confined to the $xz$-plane as 
\begin{gather}
\vb{x}_0 = (r_0, b)\\
\dot{\vb{x}}_0 = (-v_0, 0).
\end{gather}
We fix $r_0 = 100\,\mathrm{pc}$,
and use model parameters
$\rho = 400 \,\Msun\,\mathrm{pc}^{-3}$,
$\sigma=200\,\mathrm{km\,s^{-1}}$,
$M_\bullet=10^8\,\Msun$,
and
$e_\mathrm{s}=0.85$,
which match fairly well with the central properties of the merging galaxy when
the SMBHs are becoming bound in the $e_0=0.99$ merger orbit simulations.
\drafting{(Check the params)}

>>>>

After specifying the initial position and velocity, as well as the other parameters of the model,
the equation of motion can be solved using standard numerical integration techniques.
We use the error controlled 8th order Runge--Kutta method DOP853 included in the SciPy library \citep{virtanen2020}
for this purpose.
To mimic the nearly linearly plunging orbit in the high-eccentricity merger orbit simulations,
we specify the initial conditions confined to the $xz$-plane as 
\begin{gather}
\vb{x}_0 = (r_0, b)\\
\dot{\vb{x}}_0 = (-v_0, 0).
\end{gather}
We fix $r_0 = 100\,\mathrm{pc}$,
and use model parameters
$\rho = 400 \,\Msun\,\mathrm{pc}^{-3}$,
$\sigma=200\,\mathrm{km\,s^{-1}}$,
$M_\bullet=10^8\,\Msun$,
and
$e_\mathrm{s}=0.9$,
which are similar to the central properties of the merging galaxy when
the SMBHs are becoming bound in the $e_0=0.90$ and $e_0=0.99$ merger orbit simulations.
\drafting{(Check the params)}
